Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-02 sections/eval.tex
Update on Overleaf.

\subsubsection{Standardizing \& Heuristic Optimization}
Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting. The semantically important requirements of this standard form are 1) all inputs (i.e. tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator. The former allows the interpreter to identify the loop order for each kernel. The second determines the output format for each intermediate. The last one guarantees that the innermost expression can be computed as a single kernel. To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes. In the future,

\subsubsection{Finch Interpreter} 
Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process. First, the output format is identified by unpacking the outer $\finchreformat$ statement. Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to convert the pointwise expression into a Finch expression. At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor. Lastly, the concordant loop order is identified and instantiated. The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.

>>>>

\subsubsection{Standardizing \& Heuristic Optimization}
Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting. The semantically important requirements of this standard form are 1) all inputs (i.e. tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator. The former allows the interpreter to identify the loop order for each kernel. The second determines the output format for each intermediate. The last one guarantees that the innermost expression can be computed as a single kernel. To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes. In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.

\subsubsection{Finch Interpreter} 
Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process. The output format is identified by unpacking the outer $\finchreformat$ statement. Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to process the pointwise expression into a Finch expression. At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor. Lastly, the concordant loop order is identified and instantiated. The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.
